Binning Strategies and Related Loss for Large Data

### 2015-12-15 versionForJCGS/PaperJCGS.Rnw
Version for jcgs

\documentclass[letterpaper,10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amsthm,amssymb}
\usepackage{fullpage}
\usepackage{graphicx,float,wrapfig,subfig,tabularx,ulem}
\graphicspath{{figure/}}
%\usepackage{csquotes}
\usepackage{color}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{mathrsfs}

@@

%opening
\title{Binning Strategies and Related Loss for Large Data}
\author{Karsten Maurer$^1$, Susan VanderPlas$^1$, Heike Hofmann$^{1,2}$\\$^1$Department of Statistics, $^2$Human Computer Interaction\\Iowa State University}

@@

\begin{abstract}
Dealing with the data deluge of the Big Data Age is both exciting and challenging. The demands of large data require us to re-think strategies of visualizing data. Plots employing binning methods have been suggested in the past as viable alternative to standard plots based on raw data, as the resulting area plots tend to not be affected by increases in data as much. This comes with the price of loss of information inherent to any binning scheme. In this paper we discuss properties of two commonly used binning algorithms. We define loss of information in the specific setting of two dimensional displays, provide readily applicable tools for loss evaluation and discuss the two binning schemes with respect to their loss in the framework of a simulation and two case studies.
\end{abstract}

\section{Introduction}
\label{Intro}
% \hh{General: write out any positive integer less than or equal to ten out in words, even if there is the word `unit' behind it. So it's one unit, and not 1 unit.}
% 
% \hh{
% \begin{itemize}
% \item Additional literature: binning papers: standard binning, e.g.: Karl Pearson (1895) ``Contributions to the Mathematical Theory of Evolution. II. Skew Variation in Homogeneous Material"\\
% David W. Scott (1979). "On optimal and data-based histograms". Biometrika 66 (3): 605–610
% 
% quantile binning: there's a lot of implementations (e.g. cut2 in package Hmisc, the rattle package, the package extracut. It would be interesting to check, if they do the same things (probably not) and mention that there is disagreement in the implementations).
% 
% As a reference I would check in 
% Scott, D. W. 1992.
% Multivariate Density Estimation: Theory, Practice, and Visualization. New York: Wiley.
% 
% \item make sure to verify that the binning algorithms agree with the literature (and check that the code is doing the same thing as described in the paper).
% You can include a bit of discussion on the problem that histograms are over-parameterized - it's a known problem, but I can't find any paper on it at the moment.
% 
% \end{itemize}
% }
% 
% \ktm{I have read through the sections of Pearson (1895), Scott (1979) and Scott (1992) that pertained to binning. I will work on integrating them into the lit review next.  Scott talked much about the optimal choice of bin width based on MSE but there was not much talk about what to do with points at the boundaries of bins (he effectively assumed infinite precision in the measurement process to avoid it). 
% 
% I will need to search more to find papers about the binning algorithms and the boundaries they set.  The \texttt{cut} function allows for intervals that are closed on either the left or right and also allows the points on the outermost open bound to be included in the last bin.  
% 
% The \texttt{cut2} function allows the same type of quantile binning as discussed in Scott (1992).}

<<setup,echo=F,include=F,eval=T>>=
library(ggplot2)
library(gridExtra)
library(dbData)
library(reshape2)
#library(multicore)
library(devtools)
library(xtable)
library(RColorBrewer)
library(hexbin)
library(MASS)
library(scales)
library(plyr)
#source in binning functions
source("BinningLossFunctions.R")

connect <- dbConnect(dbDriver("MySQL"), host="mysql2.stat.iastate.edu", 
                     port=3306, user="dbaccess", dbname="baseball")
pitch <- new("dataDB", co=connect, table="Pitching")

d1 <- dbData(pitch, vars=c( "G", "SO")) # Games, Strike Outs

xtd.data <- function(x) {
# replicate each line as often as given by the last entry
  k <- dim(x)[2]

  res <- data.frame(sapply(1:(k-1), function(var) {
			return(rep(x[,var], x[,k]))
		}
	))
	names(res) <- names(x)[1:(k-1)]
	return(res)
}
d1.ext <- xtd.data(d1)
dbDisconnect(connect)
@

@@

Technological advances have facilitated collection and dissemination of large data more and more as records are digitized and our lives are increasingly lived online. According to an EMC report in 2014 ``the digital universe is doubling in size every two years and will multiply 10-fold between 2013 and 2020 - from 4.4 trillion gigabytes to 44 trillion gigabytes" (\url{http://www.emc.com/about/news/press/2014/20140409-01.htm}). This ``Data Deluge" of the Big Data Age (NY Times, Feb 2012) poses exciting challenges to data scientists everywhere: ``It's a revolution $\dots$ The march of quantification, made possible by enormous new sources of data, will sweep through academia, business and government. There is no area that is going to be untouched"-- Gary King, Harvard Institute.

Data sets with millions of records and thousands of variables are not uncommon. \citet{Friedman97} proposed in his paper on data mining and statistics that ``Every time the amount of data increases by a factor of ten, we should totally rethink how we analyze it". \citet{jacobs2009} echoed the sentiment, stating that ``big data should be defined at any point in time as \textit{data whose size forces us to look beyond the tried-and-true methods that are prevalent at that time}". The same holds true for visualizations. With a 100-1000 fold increase in the amount of data, the utility of some of our most used graphical tools, such as scatterplots, deteriorates quickly \citep{gold}.

Area plots, such as histograms, do not tend to be as affected by increases in the amount of data as plots that utilize raw data. By using binning strategies and the principles for displaying information in area plots, scatterplots can again become useful instruments for large data settings \citep{gold}.

In this paper we describe first the problem scatterplots are exposed to in large-data situations. We discuss different binning algorithms use the construct binned scatterplots and the  {\it loss of information} inherent to binning. \ktm{We will then explore the effects of binning specification on the properties of binned scatterplots through simulation and real-data case studies.} We conclude with several practical suggestions for binning specifications for creating binned scatterplots that have desirable visual properties.

\hh{XXX `binning properties' is too vague. Put in a bit more detail.}

@@

In the case of modestly sized data, scatterplots are great tools for showing relationships in two dimensions. With large data, scatterplots suffer from over-plotting of points, which masks relevant structure. Figure~\ref{fig:scatter-alpha} shows an example taken from baseball statistics. The scatterplot shows 139 seasons (from the years of 1871 -- 2009) of  pitching statistics for every baseball pitcher as published in Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/}).  The number of games played in a season is plotted versus number of strikeouts a pitcher threw over the course of a season. While the data set is only medium sized with \Sexpr{dim(d1.ext)[1]} observations, it already shows some of the break-down patterns scatterplots experience with large data.

Figure~\ref{fig:scatter-alpha}a shows a traditional scatterplot with each observation is drawn with a filled circle. A triangular structure is apparent with some outliers at a medium number of games and high number of strikeouts; however the density within the triangular mass of points is indistinguishable. \citet{tukey} suggested the use of open circles (see Figure~\ref{fig:scatter-alpha}b) to mitigate the problem of over-plotting. Open circles make points that are close together more visually distinct; thus allowing for the perception of more density information than with filled points.  A modern alternative to open circles is alpha blending (see Figure~\ref{fig:scatter-alpha}c). Alpha blending provides more frequency information, allowing a distinction of higher-resolution density.

\begin{figure}[hbtp]
  \subfloat[Overplotted data ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{Overplotting.pdf}}
  \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{OverplottingCircles.pdf}} 
	\subfloat[Alpha blending]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{OverplottingAlpha.pdf}}

  \subfloat[Hexagonal binned scatterplot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{HexBinning.pdf}}
  \subfloat[Hexagonal binned bubble plot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{BinScatter.pdf}}

	\caption{\label{fig:scatter-alpha} Scatterplots of Games versus Strikeouts in Major League Baseball, using different strategies of dealing with the issue of over-plotting:  (a) uses standard, opaque, filled circles, (b)  uses Tukey's recommended open circles, and (c) uses filled circles with alpha blending ($\alpha$=0.05). Plots (d) and (e) show hexagonal binning strategies with frequency mapped to color and area respectively}
\end{figure}

@@

Other scatterplot adaptations have been introduced that avoid overplotting by manipulating the display of the points by distorting the locations or the scales. Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations (including ties) but use distortion of the point locations by having points repel one another to avoid overlapping points.  An extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}. Variable-binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points. This variable binning fragments the continuity of the axes into  segments on different scales and also does not deal with ties. Generalized and variable-binned scatterplots make fine data structure more visible and allow color to be reserved for a third variable instead of frequency; however, the distortion of the point locations and/or axes warp the visual display of the association between the two primary variables.

Another approach is to reduce the graphical complexity by plotting binned aggregations of the data, namely frequencies, as opposed to plotting every observation as an individual point. This has the additional advantage of reducing the size of the stored data necessary for the construction of the plot, as only the bin centers and the frequency of observations which correspond to that bin must be stored. Wickham argues for a ``bin-summarize-smooth'' procedure to be applied to the visualization of big data and he notes that simple summary functions, such as counts, scale  well with the size of data \citep{Wickham2013Bin}. Liu, Jiang and Heer employ the computational benefits of binning for their interactive big data visualization program, \textit{imMens} \citep{Liu2013imMens}.  \hh{also add \citep{gold} for binning reference - it's called `weighting' there.}

Methods commonly used to display binned variables include sunflower plots \citep{sunflowerplots}, kernel density smoothing of tonal variation and modified scatterplots \citep{martin-gold}.  Sunflower plots are scatterplots of binned data, where the symbol used for the bin increases in complexity in proportion to the number of points in that bin. Sunflower plots are particularly useful when the number of points in each bin remains reasonably small. Kernel density smoothing can be used to vary $\alpha$ or color according to a smoothed density, providing features similar to binned scatterplots or alpha blended scatterplots in a more smooth, continuous fashion. However, these estimates require careful parameter tuning, as over-smoothing may hide gaps in the data while simultaneously de-emphasizes outlying points.

Histograms are a simple example of a plot that can be built using binned aggregations of the data; in their case the bin locations and bin counts act as a set of sufficient statistics necessary to reconstruct the plot.  A natural extensions of histograms to higher dimensions is to form a tessellated grid on a two dimensional Cartesian plane using some other attribute, such as color or size (or 3D renderings, which we do not want to recommend at this point) to provide joint density information within each grid cell, known as a tile.  A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) at the bin center, similar to a two-dimensional histogram viewed from above. A \textit{bubble plot} is a binned data plot that scales the size of a filled circle in proportion to frequency.  Bubble plots were first used by William Playfair \citep{playfair, playfair2}.

Figure~\ref{fig:scatter-alpha} contains examples of a hexagonally binned scatterplot with frequency encoded as color (\ref{fig:scatter-alpha}d) and a bubble plot with frequency encoded as point size (\ref{fig:scatter-alpha}e). The hexagonal tiles and the modified scatterplot are more effective at displaying the shape of the joint density and preserving outliers than any of the scatterplots shown in Figure~\ref{fig:scatter-alpha} (a-c). The bubble plot is prone to suffer from the Hermann-grid illusion \citep{hermann:1870} especially for data binned on a grid; whereas the binned scatterplot will not on a tessellated grid.

Only in the binned scatterplot and the bubble plot the inner structure of the data becomes apparent: the joint density consists two distinct ridges following two lines with very different slopes. The lower slope corresponds to the modern average strike out rate of pitchers of just under one strike-out per game. The other line has a slope of about four times that rate. This high rate is also associated with fewer games played. Closer investigation of other, related variables reveals that this high strike-out rate corresponds mainly to historic pitchers with much shorter seasons (in 1876 only 70 games were played in a season, as opposed to 162 in 2009), and qualitatively different balls and  bats.

%\hh{... This is the place to put some background literature: include the discussion of use of binning in the literature: Martin Theus's chapter in the GOLD book, Dan Carr's hexagonal binning, sunflower plots, ... }

@@

%\hh{Main motivation: we want to investigate how much information we lose when we bin.}

%\hh{I also would like to pick up the idea of sufficient statistics for scatterplots again - as a mathematical basis for measuring the amount of information that goes into a plot (since we can't expect to ever get more out of a plot). .... I'm not sure, yet, where in the paper we should place this discussion.}

@@

Binning in dimensions $X$ and $Y$ provides us with a more condensed form of the data that ideally preserves both the joint distribution as well as the margins, while reducing the amount of information to a fraction of the original.  Binning is a two-step procedure: we first assign each observation $(x, y)$ to a bin center $(x^\ast,y^\ast)$, and in a second step we count the number of observations assigned to each unique bin center; resulting in  triples  of the form $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.

We will proceed with rectangular bins for simplicity, but other binning schemes, such as hexagonal bins \citep{scatterplots} are also common.  While hexagonal binning has been shown to have slightly better graphical properties \citep{scott1992}; rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus we can present the one-dimensional case which will easily generalize to two or more dimensions.  We will however only consider binning in up to two dimensions, $X$ and $Y$. The algorithms we discuss are immediately applicable to higher dimension, but we do not feel that the paper would benefit from a more general discussion.

@@

This is expressed mathematically using the binning function $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as 
%
\begin{eqnarray}\label{rectbin}
b_X(x_i) = \left\{\begin{array}{ll} 
  x^\ast_{1} &\text{ for all } x_i = \beta_{0} \\
  x^\ast_j & \text{ for all } x_i \in (\beta_{j-1} , \beta_j] 
  \end{array}\right.
\end{eqnarray}  
%

@@

\textit{Quantile binning} is another option that divides the range of the observations into bins each containing equal number of points. The $j^{th}$ bin intervals is $(Q_X((j-1)/J),Q_X((j)/J)]$, where $Q_X(p)$ is the the $p^{th}$ empirical quantile using the inverse empirical distribution function. Note that this binning approach is \textit{not} desirable for visualizing density patterns, as it effectively balances the frequency counts in all bins; it does however have desirable properties for creating discernible groups in visualization using discrete color palettes.

@@

\begin{table}[hbtp]
\centering
\begin{tabular}{lll} \hline
 & Bin Boundaries & Bin Centers \\ 
 \hline  
General &  $ \{\beta_j \text{ }|\text{ } \beta_j > \beta_{j-1} \} $ & $\{x_j^\ast \text{ }|\text{ } x_j^\ast = (\beta_{j-1}+ \beta_j)/2 \}$ \\
%-------------
Standard \hspace{0.5cm} & $ \{\beta_j \text{ }|\text{ } \beta_j = \beta_{j-1} + \omega_X \} $\hspace{0.5cm} & $\{x_j^\ast \text{ }|\text{ } x_j^\ast = \beta_{j-1} + \omega_X/2 \}$ \\
%-------------
Quantile & $ \{\beta_j \text{ }|\text{ } \beta_j = Q_X(j/J) \} $  & $\{x_j^\ast \text{ }|\text{ } x_j^\ast = Q_X((j-0.5)/J) \}$ \\
%-------------
Random &  ---  & $\{x_j^\ast \text{ }|\text{ } x_j^\ast > x_{j-1}^\ast \}$  \\
\hline
\end{tabular}
\caption{Rectangular and Random Binning Specifications}
\label{tab:rectbinning}
\end{table}
%
An alternative to the rectangular binning processes discussed above, is the random binning algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ which randomly assigns a value $x_i$ to a bin center, $x^\ast$, from a set of possible bins. In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$. The bin function assigns $x_i$ to a bin center with a probability inversely proportional to the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center. More formally,
%
\begin{eqnarray}\label{randbin1}
b^r_X(x_i) = \left\{\begin{array}{ll} 
  x^\ast_j & \text{ with probability } (x^\ast_{j+1} - x_i)/(x^\ast_{j+1} -x^\ast_{j}) \\
  x^\ast_{j+1} &\text{ with probability }  (x_i - x^\ast_{j})/(x^\ast_{j+1} -x^\ast_{j})
  \end{array}\right.
\end{eqnarray}  
%
for $x_i \in [x^\ast_{j+1}, x^\ast_{j}]$.  This probability structure is inversely proportional to the distance from each bin center.  In Table~\ref{tab:rectbinning} we note that this random binning algorithm does not specify bin boundaries; only a sequence of bin centers.  This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform distribution of bin centers.

@@

The standard and random binning algorithms are easily extendable to higher dimension. For the purposes of creating binned scatterplots we will specify extension to rectangular binning in two dimensions. In this case we wish to assign data pairs $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_k^\ast)$, with $j \in \{1,\dots,J\}$ and $k \in \{1,\dots,K\}$, where $J$ and $K$ are the number of bins in the X and Y dimensions, respectively. The two ($j$,$k$) pairs that index the bin centers can be linearized to a single index such that $\ell = j + J(k-1)$; thus making $j$ the fast running index and $k$ the slow running index.  With this linearized index for all bins we now have a set of bin centers of the form $(x_\ell^\ast,y_\ell^\ast)$, with $\ell \in \{1,\dots,\mathscr{L}\}$, where $\mathscr{L}=J\cdot K$.

>>>>

\documentclass[letterpaper,10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amsthm,amssymb}
\usepackage{fullpage}
\usepackage{graphicx,float,wrapfig,subfig,tabularx,ulem}
%\graphicspath{{figure/}}
%\usepackage{csquotes}
\usepackage{color}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{mathrsfs}

@@

%opening
\title{Binning Strategies and Related Loss for Large Data}
\author{Karsten Maurer$^1$, Susan VanderPlas$^2$, Heike Hofmann$^{2,3}$\\$^1$Department of Statistics, Miami University \\$^2$Department of Statistics, Iowa State University \\$^3$Human Computer Interaction, Iowa State University}

@@

\begin{abstract}
Dealing with the data deluge of the Big Data Age is both exciting and challenging. The demands of large data require us to re-think strategies of visualizing data. Plots employing binning methods have been suggested in the past as viable alternative to standard plots based on raw data, as the resulting area plots tend to be less affected by increases in data. This comes with the price of the loss of information inherent to any binning scheme. In this paper we discuss binning algorithms used in the construction of binned scatterplots. We define functions to quantify the loss of spatial and frequency information and discuss the effects of binning specification on loss in the framework of simulation and case studies. From this we provide several practical suggestions for binning strategies that lead to binned scatterplots with desirable visual properties. 
\end{abstract}

\section{Introduction}
\label{Intro}

<<setup4,echo=F,include=F,eval=T>>=
library(ggplot2)
library(gridExtra)
library(dbData)
library(reshape2)
#library(multicore)
library(devtools)
library(xtable)
library(RColorBrewer)
library(hexbin)
library(MASS)
library(scales)
library(plyr)
#source in binning functions
source("BinningLossFunctions.R")

# connect <- dbConnect(dbDriver("MySQL"), host="mysql2.stat.iastate.edu", 
#                      port=3306, user="dbaccess", dbname="baseball")
# pitch <- new("dataDB", co=connect, table="Pitching")
# 
# d1 <- dbData(pitch, vars=c( "G", "SO")) # Games, Strike Outs
# 
# xtd.data <- function(x) {
# # replicate each line as often as given by the last entry
#   k <- dim(x)[2]
# 
#   res <- data.frame(sapply(1:(k-1), function(var) {
#   		return(rep(x[,var], x[,k]))
# 		}
# 	))
# 	names(res) <- names(x)[1:(k-1)]
# 	return(res)
# }
# d1.ext <- xtd.data(d1)
# dbDisconnect(connect)
# save(d1, d1.ext, file="pitching.Rdata")
load("pitching.Rdata")
@

@@

Technological advances have facilitated collection and dissemination of large data as records are digitized and our lives are increasingly lived online. According to an EMC report in 2014 ``the digital universe is doubling in size every two years and will multiply 10-fold between 2013 and 2020 - from 4.4 trillion gigabytes to 44 trillion gigabytes" (\url{http://www.emc.com/about/news/press/2014/20140409-01.htm}). This ``Data Deluge" of the Big Data Age (NY Times, Feb 2012) poses exciting challenges to data scientists everywhere: ``It's a revolution $\dots$ The march of quantification, made possible by enormous new sources of data, will sweep through academia, business and government. There is no area that is going to be untouched"-- Gary King, Harvard Institute.

Data sets with millions of records and thousands of variables are not uncommon. \citet{Friedman97} proposed in his paper on data mining and statistics that ``Every time the amount of data increases by a factor of ten, we should totally rethink how we analyze it". \citet{jacobs2009} echoed the sentiment, stating that ``big data should be defined at any point in time as \textit{data whose size forces us to look beyond the tried-and-true methods that are prevalent at that time}". The same holds true for visualizations. With a 100-1000 fold increase in the amount of data, the utility of some of our most commonly used graphical tools, such as scatterplots, deteriorates quickly \citep{gold}.

Area plots, such as histograms, do not tend to be as affected by increases in the amount of data because they display aggregations instead of raw data. By using binning strategies and the principles for displaying information in area plots, scatterplots can again become useful instruments for large data settings \citep{gold}.

In this paper we describe first the inadequacy of traditional scatterplots in large-data situations. We discuss different binning algorithms use in the construction of binned scatterplots and the \textit{loss of information} inherent to binning. We will then explore the effects of binning specification on the properties of binned scatterplots through simulation and real-data case studies. We conclude with several practical suggestions for binning specifications for creating binned scatterplots that have desirable visual properties.

@@

In the case of modestly sized data, scatterplots are great tools for showing bivariate data relationships. With large data, scatterplots suffer from over-plotting of points, which masks relevant structure. Figure~\ref{fig:scatter-alpha} shows an example taken from baseball statistics. The scatterplot shows 139 seasons (from the years of 1871 -- 2009) of  pitching statistics for every baseball pitcher as published in Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/}).  The number of games played in a season is plotted against the number of strikeouts a pitcher threw over the course of a season. While the data set is only medium sized with \Sexpr{dim(d1.ext)[1]} observations, it already shows some of the break-down patterns scatterplots experience with large data.

Figure~\ref{fig:scatter-alpha}(a) shows a traditional scatterplot with each observation is drawn with a filled circle. A triangular structure is apparent with some outliers at a medium number of games and high number of strikeouts; however the density within the triangular mass of points is indistinguishable. \citet{tukey} suggested the use of open circles (see Figure~\ref{fig:scatter-alpha}(b)) to mitigate the problem of over-plotting. Open circles make points that are close together more visually distinct; thus allowing for the perception of more density information than with filled points.  A modern alternative to open circles is alpha blending (see Figure~\ref{fig:scatter-alpha}(c)). Alpha blending renders points as semi-transparent to provides more visibility of underlying points.

\begin{figure}[hbtp]
  \subfloat[Overplotted data ]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/Overplotting.pdf}}
  \subfloat[Tukey-style open circles]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/OverplottingCircles.pdf}} 
	\subfloat[Alpha blending]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/OverplottingAlpha.pdf}}

  \subfloat[Hexagonal binned scatterplot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{./figure/HexBinning.pdf}}
  \subfloat[Hexagonal binned bubble plot]{\includegraphics[keepaspectratio=TRUE,width=.48\textwidth]{./figure/BinScatter.pdf}}

	\caption[Traditional and adapted scatterplots for games vs. strikeouts data.]{\label{fig:scatter-alpha} Scatterplots of games versus strikeouts in major league baseball, using different strategies of dealing with the issue of over-plotting:  (a) uses standard, opaque, filled circles, (b)  uses Tukey's recommended open circles, and (c) uses filled circles with alpha blending ($\alpha$=0.05). Plots (d) and (e) show hexagonal binning strategies with frequency mapped to color and area respectively.}
\end{figure}

@@

Other scatterplot adaptations have been introduced that avoid over-plotting by manipulating the display of the points by distorting the locations or the scales. Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations, including those sharing identical coordinates, and use distortion of the point locations by having points repel one another to avoid overlapping. An extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}. Variable-binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points. This variable binning fragments the continuity of the axes into segments on different scales and also does not deal with points at identical coordinates. Generalized and variable-binned scatterplots make fine data structure more visible and allow color to be reserved for a third variable instead of frequency; however, the distortion of the point locations and/or axes warp the visual display of the association between the two primary variables.

Another approach is to reduce the graphical complexity by plotting binned aggregations of the data, namely frequencies, as opposed to plotting every observation as an individual point. This has the additional advantage of reducing the size of the stored data necessary for the construction of the plot, as only the bin centers and the bin frequencies must be stored. Wickham argues for a ``bin-summarize-smooth'' procedure to be applied to the visualization of big data and he notes that simple summary functions, such as counts, scale  well with the size of data \citep{Wickham2013Bin}. Liu, Jiang and Heer employ the computational benefits of binning for their interactive big data visualization program, \textit{imMens} \citep{Liu2013imMens}.

Methods commonly used to display binned variables include sunflower plots \citep{sunflowerplots}, kernel density smoothing of tonal variation and binned scatterplots \citep{martin-gold}. Sunflower plots are scatterplots of binned data, where the symbol used for the bin increases in complexity in proportion to the number of points in that bin. Sunflower plots are particularly useful when the number of points in each bin remains reasonably small. Kernel density smoothing can be used to vary $\alpha$ or color according to a smoothed density, providing features similar to binned scatterplots or alpha blended scatterplots in a more smooth, continuous fashion. However, these estimates require careful parameter tuning, as over-smoothing may hide gaps in the data while simultaneously de-emphasizes outlying points.

Histograms are a simple example of a plot that can be built using binned aggregations of the data; in their case the bin locations and bin counts act as a set of sufficient statistics necessary to reconstruct the plot. A natural extensions of histograms to higher dimensions is to form a tessellated grid on a two dimensional Cartesian plane using some other attribute, such as color or size or 3D renderings to provide joint density information within each grid cell, known as a tile.  A \textit{bubble plot} is a binned data plot that scales the size of a filled circle in proportion to frequency.  Bubble plots were first used by William Playfair \citep{playfair, playfair2}. A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) at the bin center, similar to a two-dimensional histogram viewed from above.

Figure~\ref{fig:scatter-alpha} contains examples of a hexagonally binned scatterplot with frequency encoded as color (\ref{fig:scatter-alpha}(d)) and a bubble plot with frequency encoded as point size (\ref{fig:scatter-alpha}(e)). The hexagonally binned scatterplot and the bubble plot are more effective at displaying the shape of the joint density and preserving outliers than any of the scatterplots shown in Figure~\ref{fig:scatter-alpha}(a-c). The bubble plot is prone to suffer from the Hermann-grid illusion \citep{hermann:1870}, where the white spaces between circles on the evenly spaced grid appear shaded due to an optical illusion; whereas this will not occur for a binned scatterplot on a tessellated grid.

The inner structure of the baseball data is only apparent in the binned scatterplot and the bubble plot. The joint density consists two distinct ridges following two lines with very different slopes. The lower slope corresponds to the modern average strike out rate of pitchers of just under one strike-out per game. The other line has a slope of about four times that rate. This high rate is also associated with fewer games played. Closer investigation of other, related variables reveals that this high strike-out rate corresponds mainly to historic pitchers with much shorter seasons (in 1876 only 70 games were played in a season, as opposed to 162 in 2009), and qualitatively different balls and bats.

@@

@@

Binning in dimensions $X$ and $Y$ provides us with a more condensed form of the data that ideally preserves both the joint distribution as well as the margins, while reducing the amount of information to a fraction of the original.  Binning is a two-step procedure: we first assign each observation $(x, y)$ to a bin center $(x^\ast,y^\ast)$, and in a second step we count the number of observations assigned to each unique bin center; resulting in reduced data triples of the form $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.

We will proceed with rectangular bins for simplicity, but other binning schemes, such as hexagonal bins \citep{Carr1987} are also common.  While hexagonal binning has been shown to have slightly better graphical properties \citep{scott1992}; rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus we can present the one-dimensional case which will easily generalize to two or more dimensions.  We will however only consider binning in up to two dimensions, $X$ and $Y$. The algorithms we discuss are immediately applicable to higher dimension, but we do not feel that the paper would benefit from a more general discussion.

@@

This is expressed mathematically using the binning function $b_X(.) : x_i \rightarrow x^\ast_j$ defined as 
%
\begin{eqnarray}\label{rectbin}
b_X(x_i) = \left\{\begin{array}{ll} 
  x^\ast_{1} &\text{ for all } x_i = \beta_{0} \\
  x^\ast_j & \text{ for all } x_i \in (\beta_{j-1} , \beta_j] 
  \end{array}\right.
\end{eqnarray}  
%

@@

\textit{Quantile binning} is another option that divides the range of the observations into bins each containing an equal number of points. The $j^{th}$ bin interval takes the form $(Q_X((j-1)/J),Q_X((j)/J)]$, where $Q_X(p)$ is the the $p^{th}$ empirical quantile using the inverse empirical distribution function. Note that this binning approach is \textit{not} desirable for spatially visualizing density patterns, as it effectively balances the frequency counts in all bins; it does however have desirable properties for binned scatterplots that employ a second stage of binning to create discrete shade scheme for displaying grouped bin frequencies, which will be discussed in Section~\ref{FreqLoss}.

@@

\begin{table}[hbtp]
\centering
\caption{Rectangular and Random Binning Specifications}
\begin{tabular}{lll} \hline
 & Bin Boundaries & Bin Centers \\ 
 \hline  
General &  $ \{\beta_j \text{ }|\text{ } \beta_j > \beta_{j-1} \} $ & $\{x_j^\ast \text{ }|\text{ } x_j^\ast = (\beta_{j-1}+ \beta_j)/2 \}$ \\
%-------------
Standard \hspace{0.5cm} & $ \{\beta_j \text{ }|\text{ } \beta_j = \beta_{j-1} + \omega_X \} $\hspace{0.5cm} & $\{x_j^\ast \text{ }|\text{ } x_j^\ast = \beta_{j-1} + \omega_X/2 \}$ \\
%-------------
Quantile & $ \{\beta_j \text{ }|\text{ } \beta_j = Q_X(j/J) \} $  & $\{x_j^\ast \text{ }|\text{ } x_j^\ast = Q_X((j-0.5)/J) \}$ \\
%-------------
Random &  ---  & $\{x_j^\ast \text{ }|\text{ } x_j^\ast > x_{j-1}^\ast \}$  \\
\hline
\end{tabular}
\label{tab:rectbinning}
\end{table}
%
An alternative to the rectangular binning processes discussed above, is the random binning algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ to randomly assigns an observation, $x_i$, to a bin center, $x^\ast$, from a set of possible bins. In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$. The bin function assigns $x_i$ to a bin center with a probability inversely proportional to the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center. More formally,
%
\begin{eqnarray}\label{randbin1}
b^r_X(x_i) = \left\{\begin{array}{ll} 
  x^\ast_j & \text{ with probability } (x^\ast_{j+1} - x_i)/(x^\ast_{j+1} -x^\ast_{j}) \\
  x^\ast_{j+1} &\text{ with probability }  (x_i - x^\ast_{j})/(x^\ast_{j+1} -x^\ast_{j})
  \end{array}\right.
\end{eqnarray}  
%
for $x_i \in [x^\ast_{j+1}, x^\ast_{j}]$. In Table~\ref{tab:rectbinning} we note that this random binning algorithm does not specify bin boundaries; only a sequence of bin centers.  This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform distribution of bin centers.

@@

The standard and random binning algorithms are easily extendable to higher dimension. For the purposes of creating binned scatterplots we will specify extension to rectangular binning in two dimensions. In this case we wish to assign data pairs $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_k^\ast)$, with $j \in \{1,\dots,J\}$ and $k \in \{1,\dots,K\}$,where $J$ and $K$ are the number of bins in the X and Y dimensions, respectively.
The ($j$,$k$) pairs that index the bin centers can be linearized to a single index such that $\ell = j + J(k-1)$; thus making $j$ the fast running index and $k$ the slow running index.
With this linearized index for all bins we now have a set of bin centers of the form $(x_\ell^\ast,y_\ell^\ast)$, with $\ell \in \{1,\dots,\mathscr{L}\}$, where $\mathscr{L}=J\cdot K$.

The standard rectangular binning function $b(.) : (x_i,y_i) \rightarrow (x^\ast_\ell,y^\ast_\ell)$ is defined as 
%
\begin{eqnarray}\label{standrecbin1}
b(x_i,y_i) = (b_X(x_i),b_Y(y_i))
\end{eqnarray}
%
where $b_X(x_i)$ and $b_Y(y_i)$ are the univariate standard binning algorithms for the X and Y dimensions respectively. The random rectangular binning function, $b^r(\cdot): (x_i,y_i) \rightarrow (x^\ast_\ell,y^\ast_\ell)$ is similarly defined as
%
\begin{eqnarray}\label{randrecbin1}
b^r(x_i,y_i) = (b^r_X(x_i),b^r_Y(y_i))
\end{eqnarray}
%
where $b^r_X(x_i)$ and $b^r_Y(y_i)$ are univariate random binning algorithms for each dimension. Figure~\ref{fig:spatlossdemo2} provides an illustration of each binning process extended to a two dimensional situation.
